---
title: "Analysis of 1000 genomes genotypes and MyHeritage data with ancestry informative SNPs"
author: "Benedek Dankó"
date: 2023-09-06
output:
  html_document:
    code_folding: hide
    fig_crop: no
    toc: true
    theme: lumen
    toc_float: true
    toc_depth: 4
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.table.format = "html") 
```

***

```{r, warning=F, message=F, echo=F}
# Load libraries:
library("RCurl") 
library("tidyverse")
library("vcfR")
library("FactoMineR")
library("vcd")
library("factoextra")
library("plotly")
```

# Analysis of 1000 genomes genotypes and MyHeritage data utilizing ancestry informative SNPs from Kidd et al. and Seldin et al.

The main idea is taken from Kevin Arvai's [repo](https://github.com/arvkevi/tgviz.git) using a Python notebook. 

## Load data

```{r, warning=F, message=F, echo=F, results='hide'}
# MyHeritage raw SNP data:
myheritage_raw <- read.csv("/home/benedek_danko/helmholtz/nextcloud/myheritage/MyHeritage_raw_dna_data.csv",
                           skip=12, header = T, sep=",")
# 1K genomes project data:
one_k_samples = "ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/integrated_call_samples_v3.20130502.ALL.panel"
text_data <- getURL(one_k_samples, connecttimeout = 60)
df_samples <- read.csv(text = text_data,
               sep="\t",
               header = T) |> 
  dplyr::select(-c(X, X.1)) |> 
  column_to_rownames("sample")
# Kid et al. SNP data:
kid_vcf_file = read.vcfR("/home/benedek_danko/helmholtz/nextcloud/myheritage/Kidd.55AISNP.1kG.vcf")
```

## Plot 1K Genomes Project data using Kid et al. ancestry informative SNPs (AISNPs)

### Formatting SNP data

```{r, warning=F, message=F}
df = as.data.frame(kid_vcf_file@gt) |> 
  dplyr::select(rownames(df_samples)) |> 
  mutate(rs_id=getFIX(kid_vcf_file)[,3]) |> 
  column_to_rownames("rs_id") |> 
  t() |> 
  as.data.frame() 

df[df=="0|0"]<-0
df[df=="0|1" | df=="1|0"]<-1
df[df=="1|1"]<-3

dim(df)
nrow(df) == nrow(df_samples) 
```

### Performing dimensionality reduction with [Factor Analysis of Mixed Data (FAMD)](https://search.r-project.org/CRAN/refmans/FactoMineR/html/FAMD.html) and plotting

```{r, warning=F, message=F, fig.width=8, fig.height=7}
famd <- FAMD(df, graph=FALSE)
plot <- fviz_pca_ind(famd)
fig <- plot_ly(data = plot$data |> 
         left_join(df_samples |> 
                     rownames_to_column("name"),
                   by="name"), 
               x = ~x, 
               y = ~y,
               color=~super_pop)
fig
```

## Plot 1K Genomes Project and my own MyHeritage data using Kid et al. ancestry informative SNPs (AISNPs)

### Formatting SNP data

```{r, warning=F, message=F}
# kid_vcf_file@fix |> 
#   as.data.frame() |> 
#   dplyr::select(c(ID, REF, ALT))

myheritage_raw_sub <- myheritage_raw |> 
  filter(RSID %in% colnames(df)) |> 
  left_join(kid_vcf_file@fix |> 
              as.data.frame() |> 
              dplyr::select(c(ID, REF, ALT)),
            by=c("RSID"="ID")) |> 
  mutate(ALLELE1=substr(RESULT, 1, 1),
         ALLELE2=substr(RESULT, 2, 2),
         gt=ifelse(ALLELE1 == REF & 
                     ALLELE2 == REF,
                   0,
                   ifelse((ALLELE1 == REF & ALLELE2 == ALT) |
                            (ALLELE2 == REF & ALLELE1 == ALT),
                          1,
                          ifelse(ALLELE1 == ALT &
                                   ALLELE2 == ALT, 3,
                                 2)))) 
# table(myheritage_raw_sub$gt)
# 7 SNPs are missing from my data (could be imputed)

df_updated <- df |> 
  t() |> 
  as.data.frame() |> 
  rownames_to_column("RSID") |> 
  #filter(RSID %in% myheritage_raw_sub$RSID) |> 
  left_join(myheritage_raw_sub |> 
              dplyr::select(c(RSID, gt)),
            by="RSID") |> 
  column_to_rownames("RSID") |> 
  mutate_all(function(x) as.numeric(x)) |> 
  rowwise() %>%
  mutate(gt=ifelse(is.na(gt), 
                   which.max(table(c_across())),
                   gt), .) |> 
  mutate_all(function(x) as.character(x)) |> 
  t()
```

### Performing dimensionality reduction with [Factor Analysis of Mixed Data (FAMD)](https://search.r-project.org/CRAN/refmans/FactoMineR/html/FAMD.html) and plotting

```{r, warning=F, message=F, fig.height=6.5, fig.width=8}
famd <- FAMD(df_updated, graph=FALSE)
plot <- fviz_pca_ind(famd)
plot_data <- plot$data |> 
  left_join(df_samples |> 
              rownames_to_column("name"),
            by="name") %>% 
  mutate(pop=ifelse(name=="gt", "Me", pop),
         super_pop=ifelse(name=="gt", "Me", super_pop),
         symb=ifelse(name=="gt", "Me", "Other"))

fig <- plot_ly(data = plot_data, 
               x = ~x, 
               y = ~y,
               color=~super_pop,
               symbol = ~symb, 
               symbols = c("x", "circle"),
               size=ifelse(plot_data$name=="gt", 8, 3))
fig
```

## Eucledian distance 

```{r, warning=F, message=F}
euclidian = function(
    mat1,       # Matrix with observations in COLUMNS.
    mat2=mat1   # Matrix with observations in COLUMNS.
) {
  apply(mat1, 2, function(yi) sqrt(colSums((mat2 - yi)^2)))
}

d <- euclidian(t(plot_data[plot_data$name=="gt",2:3]),
               t(plot_data[!plot_data$name=="gt",2:3])) |> 
  as.data.frame() |> 
  dplyr::rename("euc_dist"="2505") |> 
  mutate(sample=plot_data$name[plot_data$name!="gt"]) |> 
  arrange(euc_dist) |> 
  left_join(df_samples |> 
              rownames_to_column("sample"),
            by="sample")
```